---
title: "Infant MLL-rearranged AML"
repo-url: https://github.com/trichelab/trichelab.github.io/
format: 
  html:
    toc: true
    toc-depth: 3
    toc-expand: 3
    toc-title: Sections
    code-fold: true
    code-links:
      - text: Quarto (source code)
        icon: github
        href: https://github.com/trichelab/trichelab.github.io/blob/gh-pages/webR/infantMLL/infantMLL.qmd
engine: knitr
execute:
  warning: false
filters:
  - webr
webr:
  packages: ['mixtools','survival','survminer','reshape2','ggplot2','table1']
---

# Age, cell of origin, and fusions in leukemia

## MLL rearrangements

Translocations involving human chromosome 11 are a frequent occurrence in infant
leukemia. When a breakpoint is on band 23 of the long arm (11q23 abnormality), 
we often find the Mixed-Lineage-Leukemia or MLL protein (encoded by _KMT2A_) 
broken in half and stuck to some other gene. Unfortunately, "some other gene" 
can be one of hundreds of known partner genes, and nobody is sure how much the 
identity of the partner gene matters. With so many different MLL fusions arising
in patients, we might wonder there are any rules (beyond "MLL-XYZ is bad news") 
that could guide a patient toward one therapy or another. MLL rearrangements 
occur across all ages; young patients have few recurrent mutations beyond MLL.
So co-mutation doesn't help, at least not in infants. Does anything else help?

## Reading the data from COG/TARGET AML

As part of the COG/TARGET AML project, over 500 infant (younger than 3 years) 
patients with acute myeloid leukemia (AML) have been characterized by DNA and 
RNA sequencing, flow cytometry, and DNA methylation microarrays, some at both 
diagnosis and relapse. 164 of these infant patients harbor MLL translocations, 
in addition to 142 children and 56 adolescent and young adult (AYA) patients,
for a total of 362 well-characterized, MLL-rearranged pediatric AML cases.

Let's see whether we can dig anything new out of this data with basic tools.

```{r}

MLL <- read.csv(url("https://trichelab.github.io/data/MLLr.csv"), row=1)
factors <- names(MLL)[c(2, 3, 8)]
for (f in factors) MLL[[f]] <- factor(MLL[[f]])
MLL$AgeGroup <- relevel(relevel(factor(MLL$AgeGroup), 2), 3)
summary(MLL[, factors]) 

```

## Table 1: 25 years of trials

We might as well summarize the cohorts while we're here:

```{r}

library(table1) 
pvalue <- function(x, ...) {
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) p <- t.test(y ~ g)$p.value
  if (!is.numeric(y)) p <- chisq.test(table(y, g))$p.value
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

common <- names(which(table(MLL$fusion) > 5))
MLL$Fusion <- ifelse(MLL$fusion %in% common, 
                     MLL$fusion, "Other MLL fusions")
MLL$Fusion <- relevel(relevel(relevel(factor(MLL$Fusion), 6), 4), 6)
table1(~ Sex + Fusion | AgeGroup, data=MLL, extra.col=list(p=pvalue))

```

## _CD34_ and _EVI1_

Previous work has shown that high levels of the cell surface marker CD34 and
the stem cell fate gene _MECOM_ (which encodes the EVI1 protein) portend dismal 
outcomes (at least in AML and B-cell ALL patients with MLL rearrangements).

(More on this topic: [Modeling phenotypes with mixtures](https://trichelab.github.io/webR/mixtures/))

Prior to TARGET, there wasn't really a well-annotated, readily-available cohort
of MLL-rearranged AML patients who had this sort of extensive characterization
across multiple age groups. But time marches on, so now there is.


## Fitting a mixture model 

Let's fit a mixture model to the expression of these two genes in MLLr AML. 

```{r}

library(mixtools) 
set.seed(1234)
bothFit <- mvnormalmixEM(MLL[, c("MECOM", "CD34")], k=3, arbvar=FALSE)
plot.mixEM(bothFit, whichplots=2, xlab2="MECOM", ylab2="CD34")
MLL$classification <- apply(bothFit$posterior, 1, which.max)
MLL$age <- ifelse(MLL$AgeGroup == "Infant", "Infant", "Older")
MLL$CD34EVI1 <- ifelse(MLL$classification == 1, "lo", "hi") 
with(MLL, table(age, CD34EVI1))

```

Now you try it: 

```{webr-r}

library(mixtools) 
MLL <- read.csv(url("https://trichelab.github.io/data/MLLr.csv"), row=1)
set.seed(1234)
bothFit <- mvnormalmixEM(MLL[, c("MECOM", "CD34")], k=3, arbvar=FALSE)
plot.mixEM(bothFit, whichplots=2, xlab2="MECOM", ylab2="CD34")
MLL$classification <- apply(bothFit$posterior, 1, which.max)
MLL$age <- ifelse(MLL$AgeGroup == "Infant", "Infant", "Older")
MLL$CD34EVI1 <- ifelse(MLL$classification == 1, "lo", "hi") 
with(MLL, table(age, CD34EVI1))

```

## Testing out the model 

Let's see how these labels predict outcome (or don't). 

```{r}
# bothFit$mu # cluster 1 is low/low
library(survival)
library(survminer)
allSurv <- survfit(Surv(OS, OSI) ~ age + CD34EVI1, data=MLL)
ggsurvplot(allSurv, conf.int=TRUE, pval=TRUE, xlab="Overall survival (days)",
           tables.theme = theme_cleantable(), 
           palette=c("red","green","darkred","darkgreen"))

```

Perhaps it will be easier to see the differences if we stratify by age group:

```{r}
# by age group
ggsurvplot_facet(allSurv, 
                 data=MLL,
                 conf.int=TRUE, 
                 pval=TRUE, 
                 legend.labs=c("CD34/EVI1 high", "CD34 & EVI1 low"),
                 xlab="Overall survival (days)",
                 palette=c("red","green"),
                 facet.by="age")

```

Your turn (be sure to run the previous chunks so `MLL` and `bothFit` are loaded)

```{webr-r}

library(survival)
library(survminer)
allSurv <- survfit(Surv(OS, OSI) ~ age + CD34EVI1, data=MLL)
ggsurvplot_facet(allSurv, 
                 data=MLL,
                 conf.int=TRUE, 
                 pval=TRUE, 
                 legend.labs=c("CD34/EVI1 high", "CD34 & EVI1 low"),
                 xlab="Overall survival (days)",
                 palette=c("red","green"),
                 facet.by="age")
```

For the sake of downstream motif analysis, let's tabulate which cases are which:

```{r}
infantMLL <- subset(MLL, age == "Infant")
with(infantMLL, table(CD34EVI1))
# write.csv(infantMLL, file="~/Dropbox/Alex_Projects/infantMLL.csv") 
```

## _CLEC2A_

It turns out that the high-risk cases (CD34-high and/or MECOM-high) may have a 
very specific immunotherapy target that could help to reverse the poor outcomes
that these patients experience when treated with standard chemotherapy. Let's
look at the evidence for this. As usual, we see if a mixture model makes sense.

```{r}
library(mixtools)
CLEC2Afit <- normalmixEM(MLL$CLEC2A)
plot(CLEC2Afit, which=2, col2=c("darkgreen","red"), 
     xlab2="log(1 + transcripts)", main2="CLEC2A expression")
```

Looks like it. Eyeballing this, we can cut at around log(1 + transcripts) > 3.
More to the point, though: how much of this reflects CD34 and/or MECOM status?

```{r}
MLL$CLEC2Astatus <- c("lo", "hi")[ apply(CLEC2Afit$comp.prob, 1, which.max) ]
# or just
# MLL$CLEC2Astatus <- ifelse(MLL$CLEC2A > 3, "hi", "lo")
# if eyeballing it
tbl <- with(MLL, table(CLEC2Astatus, CD34EVI1))
chisq.test(tbl)

# now test by age, i.e. infants versus everyone else
apply(with(MLL, table(CLEC2Astatus, CD34EVI1, age)), 3, chisq.test)
```

It's not perfect, but it's unlikely due to chance.  We can perhaps get some 
idea of how to do better by checking the correlation of _CLEC2A_ with various
other genes (for example, _RORA_) that are often co-expressed with _CLEC2A_.

```{r}
ofInterest <- c('CLEC2A', 'CD34', 'MECOM', 'RORA', 'CD33')
pairs(MLL[, ofInterest]) 
```

What about overall correlation of these genes' expression?

```{r}
library(ggplot2)
library(reshape2)
cormat <- cor(MLL[, ofInterest])
to_plot <- melt(cormat, na.rm=TRUE)
names(to_plot)[1:2] <- c("Gene1", "Gene2")

p <- ggplot(to_plot, aes(Gene1, Gene2, fill=value)) +
 geom_tile(color="white") +
 scale_fill_gradient2(low="blue", high="red", mid="white", 
   midpoint=0, limit=c(-1,1), space="Lab", name="Correlation") +
 theme_classic() + 
 theme(axis.text.x=element_text(angle=90, vjust=1, size=12, hjust=1),
       axis.title.x=element_blank()) +
 theme(axis.text.y=element_text(size=12),
       axis.title.y=element_blank()) +
 coord_fixed()

```

So, while CD34/EVI1 status doesn't always predict CLEC2A status, it's not bad.
