---
title: "Infant MLL-rearranged AML"
repo-url: https://github.com/trichelab/trichelab.github.io/
format: 
  html:
    toc: true
    toc-depth: 3
    toc-expand: 3
    toc-title: Sections
    code-fold: true
    code-links:
      - text: Quarto (source code)
        icon: github
        href: https://github.com/trichelab/trichelab.github.io/blob/gh-pages/webR/infantMLL/infantMLL.qmd
engine: knitr
execute:
  warning: false
filters:
  - webr
webr:
  packages: ['mixtools','survival','survminer','patchwork','ggplot2']
---

## Background: age, cell of origin, and fusion genes in leukemia

Translocations involving human chromosome 11 are a frequent occurrence in infant
leukemia. When the breakpoint is on band 23 of the long arm of chromosome 11, or
more concisely, when we see an 11q23 abnormality, we usually suspect that the 
Mixed-Lineage-Leukemia or MLL protein (encoded by the _KMT2A_ gene) has been 
broken in half and stuck to some other gene. Unfortunately, "some other gene" 
can be one of hundreds of known partner genes, and nobody is sure how much the 
identity of the partner gene matters. What is certain is that 11q23 aberrations
are associated with very aggressive leukemic transformation, causing over half
of the cases seen in infants. Given the difficulty of drawing up rules for all
the different MLL fusions that can arise in a patient, we might wonder if there
are other markers that could tell us whether a patient will do well on any of
several possible therapies. MLL-rearranged leukemia occurs across all ages, but
younger patients have very few additional mutations, so we would expect them to
harbor the fewest confounding mutations (and, generally speaking, that's true). 

As part of the COG/TARGET AML project, over 500 infant (younger than 3 years) 
patients with acute myeloid leukemia (AML) have been characterized by DNA and 
RNA sequencing, flow cytometry, and DNA methylation microarrays, some at both 
diagnosis and relapse. 164 of these infant patients harbor MLL translocations, 
in addition to 142 children and 56 adolescent and young adult (AYA) patients,
for a total of 362 well-characterized, MLL-rearranged pediatric AML cases. 

```{r}

# available in the bifurcatoR package or from COG/TARGET clinical data elements
MLL <- read.csv(url("https://trichelab.github.io/data/MLL.csv"), row=1)

```

(More on this topic: [Modeling phenotypes with mixtures](https://trichelab.github.io/webR/mixtures/))

Previous work has shown that the cell surface marker CD34 and the stem cell
fate determining gene _MECOM_, which encodes the EVI1 protein, associate with 
poor prognosis, at least in MLL-rearranged AML and in B-cell ALL patients. 
Prior to TARGET, there wasn't really a well-annotated, readily-available cohort
of MLL-rearranged AML patients who had this sort of extensive characterization
across multiple age groups. But time marches on, so now there is.

```{r}

library(mixtools) 
# library(mclust)
# bothFit <- Mclust(MLL[, c("MECOM", "CD34")])
bothFit <- mvnormalmixEM(MLL[, c("MECOM", "CD34")], k=3, arbvar=FALSE)
# could be better, but at least it's fast-ish; takes about 4 seconds 
plot.mixEM(bothFit, whichplots=2, xlab2="MECOM", ylab2="CD34")
# ok

MLL$classification <- apply(bothFit$posterior, 1, which.max)
MLL$age <- ifelse(MLL$AgeGroup == "Infant", "Infant", "Child/AYA")

subset.mixEM <- function(x, i=seq_len(nrow(x))) { 
  x$x <- x$x[i, ]
  x$posterior <- x$posterior[i, ]
  return(x)
}
if (FALSE) { 
  # this throws a plot error, so skip it for now
  par(mfrow=c(1,2))
  plot(subset(bothFit, MLL$infant == "Infant"), whichplot=2, xlab2="MECOM", 
     ylab2="CD34", density=TRUE, alpha=c(0.0005, 0.005, 0.05), main2="Infants")
  plot(subset(bothFit, MLL$infant != "Infant"), whichplot=2, xlab2="MECOM", 
     ylab2="CD34", density=TRUE, alpha=c(0.0005, 0.005, 0.05), main2="Kids/AYA")
  par(mfrow=c(1,1))
}
```

Let's see how these labels predict outcome (or don't). 

```{r}
# bothFit$mu # cluster 1 is low/low
MLL$CD34EVI1 <- ifelse(MLL$classification == 1, "Low", "High") 

library(survival)
library(survminer)
allSurv <- survfit(Surv(OS, OSI) ~ age + CD34EVI1, data=MLL)
ggsurvplot(allSurv, conf.int=TRUE, pval=TRUE, xlab="Overall survival (days)",
           tables.theme = theme_cleantable(), 
           palette=c("red","green","darkred","darkgreen"))

# by age group
ggsurvplot_facet(allSurv, 
                 data=MLL,
                 conf.int=TRUE, 
                 pval=TRUE, 
                 legend.labs=c("CD34/EVI1 high", "CD34 & EVI1 low"),
                 xlab="Overall survival (days)",
                 palette=c("red","green"),
                 facet.by="age")

```

For the sake of downstream motif analysis, let's tabulate which cases are which:

```{r}
infantMLL <- subset(MLL, age == "Infant")
# write.csv(infantMLL, file="~/Dropbox/Alex_Projects/infantMLL.csv") 

```
