---
title: "Infant MLL-rearranged AML"
repo-url: https://github.com/trichelab/trichelab.github.io/
format: 
  html:
    toc: true
    toc-depth: 3
    toc-expand: 3
    toc-title: Sections
    code-fold: true
    code-links:
      - text: Quarto (source code)
        icon: github
        href: https://github.com/trichelab/trichelab.github.io/blob/gh-pages/webR/infantMLL/infantMLL.qmd
engine: knitr
execute:
  warning: false
filters:
  - webr
webr:
  packages: ['mixtools','survival','survminer','patchwork','ggplot2','table1']
---

# Age, cell of origin, and fusions in leukemia

## MLL rearrangements

Translocations involving human chromosome 11 are a frequent occurrence in infant
leukemia. When a breakpoint is on band 23 of the long arm (11q23 abnormality), 
we often find the Mixed-Lineage-Leukemia or MLL protein (encoded by _KMT2A_) 
broken in half and stuck to some other gene. Unfortunately, "some other gene" 
can be one of hundreds of known partner genes, and nobody is sure how much the 
identity of the partner gene matters. What is sure is that 11q23 breaks cause 
aggressive leukemic transformation, and are responsible for well over half of 
the cases seen in infants. This might be due to a fetal cell of origin, or the
fusion partner, or where the leukemia cells get stuck, but it's hard to say.

Given the many different MLL fusions that can arise in a patient, we may wonder
if there are any "rules" that can indicate if a patient will benefit from any of
several possible therapies. MLL-rearranged leukemia occurs across all ages, but
younger patients have very few recurrent co-mutations besides MLL itself. So it
seems that co-mutation status isn't very helpful, at least not in infant cases.

## Reading the data from COG/TARGET AML

As part of the COG/TARGET AML project, over 500 infant (younger than 3 years) 
patients with acute myeloid leukemia (AML) have been characterized by DNA and 
RNA sequencing, flow cytometry, and DNA methylation microarrays, some at both 
diagnosis and relapse. 164 of these infant patients harbor MLL translocations, 
in addition to 142 children and 56 adolescent and young adult (AYA) patients,
for a total of 362 well-characterized, MLL-rearranged pediatric AML cases.

Let's see whether we can dig anything new out of this data with basic tools.

```{r}

# available in the bifurcatoR package or from COG/TARGET clinical data elements
MLL <- read.csv(url("https://trichelab.github.io/data/MLL.csv"), row=1)
factors <- names(MLL)[c(2, 3, 8)]
for (f in factors) MLL[[f]] <- factor(MLL[[f]])
MLL$AgeGroup <- relevel(relevel(factor(MLL$AgeGroup), 2), 3)
summary(MLL[, factors]) 

```

## Table 1: 25 years of trials

We might as well summarize the cohort while we're here:

```{r}

library(table1) 
pvalue <- function(x, ...) {
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) p <- t.test(y ~ g)$p.value
  if (!is.numeric(y)) p <- chisq.test(table(y, g))$p.value
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

MLL <- read.csv(url("https://trichelab.github.io/data/MLL.csv"), row=1)
MLL$AgeGroup <- relevel(relevel(factor(MLL$AgeGroup), 2), 3)
common <- names(which(table(MLL$fusion) > 5))
MLL$Fusion <- ifelse(MLL$fusion %in% common, 
                     MLL$fusion, "Other MLL fusions")
MLL$Fusion <- relevel(relevel(relevel(factor(MLL$Fusion), 6), 4), 6)
table1(~ Sex + Fusion | AgeGroup, data=MLL, extra.col=list(p=pvalue))

```

## CD34 and EVI1

Previous work has shown that high levels of the cell surface marker CD34 and
the stem cell fate gene _MECOM_ (which encodes the EVI1 protein) portend dismal 
outcomes (at least in AML and B-cell ALL patients with MLL rearrangements).

(More on this topic: [Modeling phenotypes with mixtures](https://trichelab.github.io/webR/mixtures/))

Prior to TARGET, there wasn't really a well-annotated, readily-available cohort
of MLL-rearranged AML patients who had this sort of extensive characterization
across multiple age groups. But time marches on, so now there is.


## Fitting a mixture model 

Let's fit a mixture model to the expression of these two genes in MLLr AML. 

```{r}

library(mixtools) 
set.seed(12345)
bothFit <- mvnormalmixEM(MLL[, c("MECOM", "CD34")], k=3, arbvar=FALSE)
plot.mixEM(bothFit, whichplots=2, xlab2="MECOM", ylab2="CD34")
MLL$classification <- apply(bothFit$posterior, 1, which.max)
MLL$age <- ifelse(MLL$AgeGroup == "Infant", "Infant", "Older")
MLL$CD34EVI1 <- ifelse(MLL$classification == 1, "lo", "hi") 
with(MLL, table(age, CD34EVI1))

```

Now you try it: 

```{webr-r}

library(mixtools) 
MLL <- read.csv(url("https://trichelab.github.io/data/MLL.csv"), row=1)
set.seed(12345)
bothFit <- mvnormalmixEM(MLL[, c("MECOM", "CD34")], k=3, arbvar=FALSE)
plot.mixEM(bothFit, whichplots=2, xlab2="MECOM", ylab2="CD34")
MLL$classification <- apply(bothFit$posterior, 1, which.max)
MLL$age <- ifelse(MLL$AgeGroup == "Infant", "Infant", "Older")
MLL$CD34EVI1 <- ifelse(MLL$classification == 1, "lo", "hi") 
with(MLL, table(age, CD34EVI1))

```

## Testing out the model 

Let's see how these labels predict outcome (or don't). 

```{r}
# bothFit$mu # cluster 1 is low/low
library(survival)
library(survminer)
allSurv <- survfit(Surv(OS, OSI) ~ age + CD34EVI1, data=MLL)
ggsurvplot(allSurv, conf.int=TRUE, pval=TRUE, xlab="Overall survival (days)",
           tables.theme = theme_cleantable(), 
           palette=c("red","green","darkred","darkgreen"))

```

Perhaps it will be easier to see the differences if we stratify by age group:

```{r}
# by age group
ggsurvplot_facet(allSurv, 
                 data=MLL,
                 conf.int=TRUE, 
                 pval=TRUE, 
                 legend.labs=c("CD34/EVI1 high", "CD34 & EVI1 low"),
                 xlab="Overall survival (days)",
                 palette=c("red","green"),
                 facet.by="age")

```

Your turn (be sure to run the previous chunks so MLL and bothFit are loaded)

```{webr-r}

library(survival)
library(survminer)
allSurv <- survfit(Surv(OS, OSI) ~ age + CD34EVI1, data=MLL)
ggsurvplot_facet(allSurv, 
                 data=MLL,
                 conf.int=TRUE, 
                 pval=TRUE, 
                 legend.labs=c("CD34/EVI1 high", "CD34 & EVI1 low"),
                 xlab="Overall survival (days)",
                 palette=c("red","green"),
                 facet.by="age")
```

For the sake of downstream motif analysis, let's tabulate which cases are which:

```{r}
infantMLL <- subset(MLL, age == "Infant")
with(infantMLL, table(CD34EVI1))
# write.csv(infantMLL, file="~/Dropbox/Alex_Projects/infantMLL.csv") 
```
