---
title: "archRiSEE demo"
author: "Tim Triche, Jr."
format:
  html:
    embed-resources: true
    code-fold: false
    number-sections: true
toc: true
toc-title: Table of contents
toc-depth: 4
toc-location: left
execute: 
  eval: false
  echo: true
output:
  html_document:
  output-file: archRiSEEdemo.html
---

if needed, or to update:

```{r}
# Lines 2
#
```

BiocManager::install("trichelab/archRiSEE")

```{r}
# Lines 4-7
#
library(archRiSEE)
date()
```

[1] "Thu Jan 22 18:37:14 2026"

```{r}
# Lines 10
packageVersion("archRiSEE")
```

[1] '0.2.22'

```{r}
# Lines 13
system.time(ewsSub <- readRDS("ewsSub.rds"))
```

user  system elapsed 
145.217   4.084 149.362 

```{r}
# Lines 17
show(ewsSub)
```

class: SingleCellExperiment 
dim: 6062095 6000 
metadata(2): NMF NMF_UMAP
assays(2): TileMatrix logcounts
rownames(6062095): chr1:0-499 chr1:500-999 ... chrX:156040000-156040499
chrX:156040500-156040999
rowData names(33): idx assay ... nmf29 nmf30
colnames(6000): CHLA10_Bulk_Edit_JAAM7_WHLT4#AAAGATGTCCTATCAT-1
CHLA10_Bulk_Edit_JAAM7_WHLT4#AAAGGATAGTCAGCCC-1 ...
TC71_Subclone_Preimplantation_JAAM10#TTAACGGGTTTGAAGA-1
TC71_Subclone_Preimplantation_JAAM10#TTCGATTCATGGCCCA-1
colData names(61): Sample TSSEnrichment ... nmf29 nmf30
reducedDimNames(4): UMAP LSI NMF NMF_UMAP
mainExpName: FragmentCounts
altExpNames(3): GeneScores mtVariants CNA

start igvR

```{r}
# Lines 35-37
library(igvR)
igv <- igvR()
setGenome(igv, "hg38")
```

add chromatin state maps for a few Ewing sarcoma patients to spruce it up

```{r}
# Lines 40-52
library(rtracklayer) 
displayTrack(igv,
             GRangesAnnotationTrack("EWS119_chromHMM", 
                                    displayMode="COLLAPSED", 
                                    import("EWS119.chromHMM.hg38.bed.gz")))
displayTrack(igv,
             GRangesAnnotationTrack("EWS120_chromHMM",
                                    displayMode="COLLAPSED", 
                                    import("EWS120.chromHMM.hg38.bed.gz")))
displayTrack(igv,
             GRangesAnnotationTrack("EWS121_chromHMM",
                                    displayMode="COLLAPSED", 
                                    import("EWS121.chromHMM.hg38.bed.gz")))
```

check and see if these lifted-over chromatin state maps are complete crap

```{r}
# Lines 55
showGenomicRegion(igv, "CD99") 
```

actually this is a bit better. they are not complete and total garbage

```{r}
# Lines 58
showGenomicRegion(igv, "chrX:2,689,731-2,696,233") 
```

add some NMF tracks from the EWS mtscATAC model 

```{r}
# Lines 61
for (n in c("nmf1", "nmf3", "nmf5", "nmf10")) addIgvTrack(ewsSub, n, igv)
```

navigate to some gene you care about...

```{r}
# Lines 64
showGenomicRegion(igv, "NKX2-2") 
```

Try zooming out a bit to really see the differences around NKX2-2 ;-) 
igv.js can run in the background while iSEE runs in the foreground 

compare with CD99 from previously 

```{r}
# Lines 69
showGenomicRegion(igv, "chrX:2,689,731-2,696,233")
```

notice that compared to NKX2-2 or similar,
there's not much information to distinguish the NMF factors around CD99
which is kind of expected given that it's expressed in most EWS cells

look at some variants with a cutoff (can also just look at raw freqs)

```{r}
# Lines 75-83
grouping <- c("cellLine", "edited")
loci <- c(G11922A="11922G>A",
          C11925T="11925C>T",
          C11946T="11946C>T")
cutoff <- 0.05 # five percent
res <- variantFreq(ewsSub, grouping=grouping, loci=loci, cutoff=cutoff) 
names(res)[seq_along(loci)] <- names(loci) # rename to avoid annoyances
with(res, xtabs(G11922A ~ edited + cellLine))
```

|            | *CHLA10* | *TC32* | *TC71* |
|-----------:|---------:|-------:|-------:|
|   *edited* |     1086 |   1040 |   1335 |
| *unedited* |        1 |      0 |     15 |

```{r}
# Lines 89
with(res, xtabs(C11925T ~ edited + cellLine))
```

|            | *CHLA10* | *TC32* | *TC71* |
|-----------:|---------:|-------:|-------:|
|   *edited* |        3 |    414 |    846 |
| *unedited* |        0 |      1 |      1 |

```{r}
# Lines 95
with(res, xtabs(C11946T ~ edited + cellLine))
```

|            | *CHLA10* | *TC32* | *TC71* |
|-----------:|---------:|-------:|-------:|
|   *edited* |        0 |      0 |      2 |
| *unedited* |        0 |      2 |      0 |

toss these in to play with them in iSEE (e.g. facet the middle panel)

```{r}
# Lines 102-104
ewsSub$G11922A <- assay(altExp(ewsSub, "mtVariants"))['11922G>A', ]
ewsSub$C11925T <- assay(altExp(ewsSub, "mtVariants"))['11925C>T', ]
ewsSub$C11946T <- assay(altExp(ewsSub, "mtVariants"))['11946C>T', ]
```

if this looks suspiciously like `res`, that's because it is

```{r}
# Lines 107
iSEEarchR(ewsSub)
```

Listening on http://127.0.0.1:7657

Suggested exercises:

* modify the middle box to model and facet MT variants by condition
* use dynamic variable selection so middle/right boxes "listen" to UMAP (left)
* poke around at subclusters of cells or mask off entire cell lines to explore

```{r}
# Lines 115
#
```

(Ctrl-C will stop the iSEE task and hand control back to R)

```{r}
# Lines 117
#
```

things I'd like to add:

```{r}
# Lines 120
#
```

* per-pseudobulk or per-cluster CNA tracks (like with the nmf tracks above)
* on-the-fly reclustering or per-cluster views in iSEE (not unlike item 1) 

```{r}
# Lines 123
#
```

Convert this into a quarto document so I can dump it onto trichelab.github.io

```{r}
# Lines 126-128
library(quartify)
rtoqmd("archRiSEEdemo.R", "archRiSEEdemo.qmd", 
       title="archRiSEE demo", author="Tim Triche, Jr.")
```

✔ Quarto markdown file created: archRiSEEdemo.qmd
Rendering Quarto document to HTML...
✔ HTML file created: archRiSEEdemo.html
