---
title: "My title"
author: "Your name"
format:
  html:
    embed-resources: true
    code-fold: false
    number-sections: true
toc: true
toc-title: Table of contents
toc-depth: 4
toc-location: left
execute: 
  eval: false
  echo: true
output:
  html_document:
  output-file: archRiSEEdemo.html
---

if needed, or to update:

```{r}
# Lines 2
#
```

BiocManager::install("trichelab/archRiSEE")

```{r}
# Lines 4-6
#
library(archRiSEE)
packageVersion("archRiSEE")
```

[1] '0.2.6'

```{r}
# Lines 9
(ewsSub <- readRDS("ewsSub.rds"))
```

...time passes...

```{r}
# Lines 11
#
```

class: SingleCellExperiment 
dim: 6062095 6000 
metadata(2): NMF NMF_UMAP
assays(2): TileMatrix logcounts
rownames(6062095): chr1:0-499 chr1:500-999 ... chrX:156040000-156040499
chrX:156040500-156040999
rowData names(33): idx assay ... nmf29 nmf30
colnames(6000): CHLA10_Bulk_Edit_JAAM7_WHLT4#AAAGATGTCCTATCAT-1
CHLA10_Bulk_Edit_JAAM7_WHLT4#AAAGGATAGTCAGCCC-1 ...
TC71_Subclone_Preimplantation_JAAM10#TTAACGGGTTTGAAGA-1
TC71_Subclone_Preimplantation_JAAM10#TTCGATTCATGGCCCA-1
colData names(61): Sample TSSEnrichment ... nmf29 nmf30
reducedDimNames(4): UMAP LSI NMF NMF_UMAP
mainExpName: FragmentCounts
altExpNames(3): GeneScores mtVariants CNA


start igvR

```{r}
# Lines 30-32
library(igvR)
igv <- igvR()
setGenome(igv, "hg38")
```

navigate to some gene you care about...

```{r}
# Lines 35
showGenomicRegion(igv, "NKX2-2") 
```

add some NMF tracks from the EWS mtscATAC model 

```{r}
# Lines 38
for (n in c("nmf1", "nmf3", "nmf5", "nmf10")) addIgvTrack(ewsSub, n, igv)
```

takes a second to transcribe each weight track onto igv.js
...
Try zooming out a bit to really see the differences around NKX2-2 ;-) 

```{r}
# Lines 42
#
```

igv.js can run in the background while iSEE runs in the foreground 

need to play with `mirai` to make igvR and iSEE tasks non-blocking :-/ 

```{r}
# Lines 47-49
#
iSEEarchR(ewsSub)
#
```

Adding LSI1 as colData(SCE)$LSI1
...
Adding nmf1 as colData(SCE)$NMF1
...
Loading required package: shiny

Listening on http://127.0.0.1:7657
Warning in .nextMethod(se = se) :
rowRanges (GRanges) detected - copying values to rowData

```{r}
# Lines 59
#
```

Suggested exercises:

* modify the middle box to model and facet MT variants by condition
* use dynamic variable selection so middle/right boxes "listen" to UMAP (left)
* poke around at subclusters of cells or mask off entire cell lines to explore

```{r}
# Lines 66
#
```

(Ctrl-C will stop the iSEE task and hand control back to R)

```{r}
# Lines 68
#
```

things I'd like to add:

```{r}
# Lines 71
#
```

* per-pseudobulk or per-cluster CNA tracks (like with the nmf tracks above)
* on-the-fly reclustering or per-cluster views in iSEE (not unlike item 1) 
* better handling of MT variant associations with other phenomena 

```{r}
# Lines 75
#
```

Convert this into a quarto document so I can dump it onto trichelab.github.io

```{r}
# Lines 78-81
library(quartify)
rtoqmd("archRiSEEdemo.R", "archRiSEEdemo.qmd") 
library(quarto) 
quarto_render("archRiSEEdemo.qmd") 
```

